---
title: "Untitled"
author: "Villaseñor-Derbez J.C."
date: "10/4/2019"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: "hide"
header-includes: \usepackage{mathtools}
---



```{r setup}

knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F)

suppressPackageStartupMessages({
  library(ISLR)
  library(class)
  library(skimr)
  library(tidyverse)
})


```


# Load and inspect the data

```{r}
carseats <- ISLR::Carseats

skim(carseats)
```


# Feature enginiering

We create a new feature High as the response variable following the rule: If sales are higher than the median sales, then 1. Else then 0. We will also remove some of the variables that we don't care about.


```{r}
carseats <- carseats %>% 
  mutate(High = ifelse(Sales > median(Sales), "High", "Low")) %>%
  select(-Sales, -ShelveLoc, -Urban, -US)
```


# Create testing and training sets

```{r}
# Set random seed
set.seed(42)

# Sample 50% observations as training data
train <- sample(1:nrow(carseats), 200)
carseats_train <- carseats[train,]

# The rest 50% as test data
carseats_test = carseats[-train,]
```

# Break the dataframe into untidy things

```{r}
# Training set
YTrain <- carseats_train$High
XTrain <- carseats_train %>% 
  select(-High)

# Testing set
YTest<- carseats_test$High
XTest <- carseats_test %>% 
  select(-High)
```


# Train the model

```{r}
set.seed(42)
pred_YTrain <- knn(train = XTrain, test = XTrain, cl = YTrain, k=2)
```


# Evaluate performance

```{r}
conf_train <- table(predicted = pred_YTrain, true = YTrain)
conf_train
```


```{r}
sum(diag(conf_train)/sum(conf_train))
```

# Train again, but evaluate predictions on testing set


```{r}
set.seed(42)
pred_YTest <- knn(train = XTrain, test = XTest, cl = YTrain, k=2)
```


# Evaluate performance

```{r}
conf_test <- table(predicted = pred_YTest, true = YTest)
conf_test
```


```{r}
sum(diag(conf_test)/sum(conf_test))
```

